Volunteers in the Clouds: an Architecture for Low Cost and Potentially Massive Distributed Evolutionary Computation

### 2015-08-09 volunteer-clouds-ieee-ci-si.tex
Adds Weibull fit

The main problem with Java and all the other browser-embedded
technologies is that they are not {\em universal} in the
sense than an extra component, namely, the Java virtual machine or another
virtual environment has to be installed in the browser.
Also these virtual environments have become targets of malicious attacks; % [Mario] Aqui empieza
for instance numerous vulnerabilities have been discovered in % lo que decía de la seguridad.
the proprietary Adobe Flash runtime environment and
video player \cite{ford2009analyzing,watanabe2010new}. The need for
these additional environments has been mitigated with the
adoption of the HTML5 standard \cite{anthes2012html5}.
JavaScript, however,
\cite{flanagan2006javascript} was introduced in 1997 as a
browser-embedded language and has, since then, become a set of standards
\cite{ECMA-262} that include the language and its components. Since
early on, it was adopted by the industry and also by scientists,
who used them for creating an EA on
the browser as early as 1998 \cite{jj-ppsn98}.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{img/baseline-times.png}
\caption{Comparison of time to solution for the baseline system, using a {\tt
    node.js} client, for different population sizes. Only
runs in which solution has been found are used for this graph.}
\label{fig:baseline}
\end{figure}
First, we establish the baseline performance by running the
evolutionary algorithm in a desktop client written using NodEO \cite{nodeo2014}, the basic JavaScript library we have used as base to build {\sf NodIO}. This experiment tries to
find the solution to the 40-trap function with parameters $l=4, a=1,
b=2, z=3$ and a population size equal 512. The algorithm was run until the solution (a string with all
ones) was found or five million evaluations had been performed. It
took around a minute, on average, that is $68.9694$ seconds, to
perform the fifty runs. In this experiment only 33, that is, 66\% were
successful. The experiments were repeated for population $p=1024$,
with success rate upgraded to 100\% and average duration $3.46$
seconds. Results for these two experiments
are shown in Figure \ref{fig:baseline}. %Fergu: dejar claro que es solo una isla

These two experiments establish first a baseline result and show that
it depends on the population. The volunteer computing experiments that we
%FERGU: explicar un poco más lo de "depends on the population"?, por ejemplo añadiendo
%as this is a deceptive problem, and low levels or diversity may not help to
%find the solution \cite{} o algo parecido.
will describe next do not and can not have the same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic. We will try and
prove next that that target performance can be achieved by carrying
out several experiments in different conditions. Second, they also act
as a basic performance test for the server, and also prove that the
basic server can be used for many kinds of different setups, as long
as they follow the REST protocol to deposit and obtain individuals
from the pool. It is obvious that, in this case, the server is not
really needed, since there is a single client and no interaction
between them, but there is a cost in sending and obtaining chromosomes
from the server so it is included for the sake of fairer comparison.

@@

Initial experiments were set up using the OpenShift
PaaS, which provides a free tier within which this %Ya se defnió antes PaaS
experiment could be performed. That way, and as required, the hardware
and cloud cost for this experiment were zero. Experiments were
announced through a post in Twitter and other social networks, and
results were published here \cite{DBLP:conf/gecco/GuervosG15}. For the
purpose of this paper, we repeated the announcement several times
through the months of April and then by the beginning of August. All
in all, we have the set of runs with the characteristics shown in
Table \ref{tab:runs}. In general, every experiment took several
days. No particular care was taken about the time of the announcement
or the particular wording. Every {\em experiment} consisted in running
until the solution of the 40-trap problem was found. When the correct
solution was sent to the server, the counter was updated and the pool
of solutions reset to the void set. No special care was taken to wait
until all clients had finished, thus it might happen that, in fact,
the islands running in the browser {\em spill} from one experiment to
the next. However, previous experiments have proven that the influence
of these islands in the next experiment is, in fact negligible.
%
\begin{table*}
\caption{Characteristics of the volunteer computing initial runs. \label{tab:runs}}
\begin{center}
\begin{tabular}{l|rr}
\hline
Date & \# experiments & Different IPs \\
\hline
April 4th 4/4 & 57 & 191 \\
April 24th 4/24 &  231 & 559 \\
July 31th 7/31 & 97 & 179 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{table*}
\caption{Summary of time per run, number of IPS and number of {\tt
    PUT}s per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
4/4 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 &  5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%
The table shows that every run included more than 50 experiments. The %FERGU: No debería ser mejor al revés? Cada experimento está compuesto de varias runs. De todas maneras arriba lo has definido así, así que si no quieres no lo cambies. Pero me parece más correcto que un experimento tenga runs, en vez de al revés
number of different IPs intervening in them varied from more than one
hundred to more than five hundred in the second experiment.

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}. This table shows the median number of IPs
intervening in each experiment,  median of the time needed
to finish the experiment, median number of HTTP {\tt PUT}s per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
found previously \cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same ballpark and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the last two columns, that measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: this experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
fair is fair, and we were looking for creating a framework that
allowed to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results is that although it will
happen, one out of roughly 5 or roughly 20, depending on what baseline %FERGU: one->1?
experiment you choose, it will not happen always or even a majority of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found; this is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. However, it puts a probably
unnecessary cap on the time, or number of experiments, that a user %Fergu: user no tiene "an", me lo dijo paloma. es "yuser", no "iuser", por eso no lo tiene :P
will voluntarily perform.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{img/time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots. From left to right, experiments 4/4, 4/24 and 7/31. Shades
of blue indicate how many experiments included that many unique IPs.}
\label{fig:duration}
\end{figure}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{img/time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color is related to the number of IPs
  participating in the experiment. From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:zipf:os}
\end{figure}
%
We will have to analyze experiment data a bit further to find out why
this happens and also if there is some pattern in the three sets of
experiments. An interesting question to ask, for instance, is if
adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, {\em adding} more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more so taking
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
with a bigger size to check this hypothesis; however, at this point we
have not found this convenient since there are several other issues
that will be solved. %FERGU: and that will be explained next? Algún ejemplo?

@@

It is interesting also to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, and of the order of
10 seconds. As it can be seen, it drops sharply implying there are
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP {\tt PUT}s, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of {\tt PUT}s per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and  scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, negative towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
\ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

At any rate, this also shows that a convenient way of increasing the
computing power would be increasing this minimum amount per user. This
issue led us to create the next version of the framework, which is
presented next. %Fergu: esto mola, métele más caña en otras secciones para que se vea que es una conclusión util

@@

In order to improve the number of cycles per user, we improved the
basic architecture in several different aspects. By using WebWorkers %FERGU: WebWorker separado o junto? homogeneizar
the algorithm kept running on the background even if the {\sf NodIO}
tab was not in focus. It also allowed to run several clients per
browser, as opposed to a single one as we did previously. Another
improvement involved restart of the client once the solution was
found, so that as long as the user was visiting the page it was
running an evolutionary algorithm. Finally, a small algorithmic
improvement was included: instead of having a fixed population,
population size was random and different for each client. %FERGU: mínimo y máximo?

JavaScript has been commonly used to develop client-side scripts
that handle the user interface, communicate asynchronously with servers and
update the content that is displayed \cite{flanagan2006javascript}.
Long-running scripts like those proposed in this work can interfere with the
responsiveness of the web application, because everything is handled by a
single thread. The Web Workers specification \cite{hickson2012web} defines an
API that allows Web developers to execute long-running scripts in parallel
with the scripts of the main page. Worker scripts are spawned in the
background allowing a thread-like operation using message-passing as the
communication mechanism between the worker's execution environment and the
main thread. According to the Web Workers W3C Candidate Recommendation
\cite{hickson2012web} workers are expected to be long-lived, they have a high
start-up performance cost, and a high per-instance memory cost. The Web Worker
API is implemented in current versions of both desktop and mobile web browsers.
For volunteer computing using the Web Worker API brings several advantages
over a single threaded implementation:

\begin{itemize}
\item If the browser uses a tabbed document interface, the worker script
keeps running in the background even it the user brings forward another tabbed
window.
\item Several instances of NodEO can be executed in parallel in a single web
page. %FERGU: NodEO o NodIO? (es NodEO la instancia, creo pero por si acaso :P)
\item Many implementations of the Web Worker API can use multiple CPUs for
the execution of worker scripts, in particular the execution in multiple CPUs
was tested in desktop versions of Chrome, FireFox and Safari.
\item Each NodEO instance can be restarted independently.
\end{itemize}

A sequence diagram of the W$^2$ version of {\sf NodIO} is presented
in Figure~\ref{fig:system:w2}. In this version, clients use the Web Workers
API to run two NodEO instances per browser. The sequence diagram shows the
interaction between processes in a time sequence. In this system there are
two kinds of asynchronous messages: HTTP Requests shown as bold arrows between
the client and the server where PUT and GET HTTP Methods are represented by
red and blue arrows respectively. There are also messages between workers and
the main thread in locally in the client, these are shown as thin solid arrows.
There are three processes involved:
\begin{itemize}
\item {\em nodejs} This is the server side process, implemented in the nodejs %FERGU: creo que oficialmente es Node.js, homogeneizar en el paper
framework. Web request handlers are implemented using the Express web
framework. For keeping a log of the execution the Winston library is used, and
finally the NodeIO framework is responsible for keeping the shared pool.
\item {\em Client main thread}  This process handles the user interaction
using the JQuery framework. The Chart.js library is used to display a chart of
the best fitness in the population over time. The main script also creates
worker instances and handles the user interface.
\item {\em Worker global scope} The worker execution space does not have
access to the Document Object Model (DOM) or JQuery objects in the main thread.
Asynchronous HTTP requests to the server are implemented using the XHR object. %FERGU: Qué es el objeto XHR? Alguna referencia/justificación de por qué se usa?
\end{itemize}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=4in]{img/Algorithm.pdf}
\caption{Description of the W$^2$ version of {\sf NodIO}. In this
  version, clients use Web Workers to run the evolutionary algorithm,
  with two of them per browser.}
\label{fig:system:w2}
\end{figure}

>>>>

The main problem with Java and all the other browser-embedded
technologies is that they are not {\em universal} in the
sense than an extra component, namely, the Java virtual machine or another
virtual environment has to be installed in the browser. 
Also these virtual environments have become targets of malicious attacks; % [Mario] Aqui empieza 
for instance numerous vulnerabilities have been discovered in % lo que decía de la seguridad.
the propietary Adobe Flash runtime environment and 
video player \cite{ford2009analyzing,watanabe2010new}. The need for
these additional environments has been mitigated with the 
adoption of the HTML5 standard \cite{anthes2012html5}.
JavaScript, however,
\cite{flanagan2006javascript} was introduced in 1997 as a
browser-embedded language and has, since then, become a set of standards
\cite{ECMA-262} that include the language and its components. Since
early on, it was adopted by the industry and also by scientists,
who used them for creating an EA on
the browser as early as 1998 \cite{jj-ppsn98}.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{img/baseline-times.png}
\caption{Comparison of time to solution for the baseline system, using a {\tt
    node.js} client, for different population sizes. Only
runs in which solution has been found are used for this graph.}
\label{fig:baseline}
\end{figure}
First we establish the baseline performance by running the
evolutionary algorithm in a desktop client written using NodEO \cite{nodeo2014}, the basic JavaScript library we have used as base to built {\sf NodIO}. This experiment tries to 
find the solution to the 40-trap function with parameters $l=4, a=1,
b=2, z=3$ and a population size equal 512. The algorithm was run until the solution (a string with all
ones) was found or five million evaluations had been performed. It
took around a minute, on average, that is $68.9694$ seconds, to
perform the fifty runs. In this experiment only 33, that is, 66\% were
successful. The experiments were repeated for population $p=1024$,
with success rate upgraded to 100\% and average duration $3.46$
seconds. Results for these two experiments 
are shown in Figure \ref{fig:baseline}.

These two experiments establish first a baseline result and show that
it depends on the population. The volunteer computing experiments that we
will describe next do not and can not have the same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic. We will try and
prove next that that target performance can be achieved by carrying
out several experiments in different conditions. Second, they act also
as a basic performance test for the server, and also prove that the
basic server can be used for many kind of different set-ups, as long
as they follow the REST protocolo to deposit and obtain individuals
from the pool. It is obvious that, in this case, the server is not
really needed, since there is a single client and no interaction
between them, but there is a cost in sending and obtaining chromosomes
from the server so it is included for the sake of fairer comparison.

@@

Initial experiments were set up using the OpenShift
Platform-as-a-Service, which provides a free tier within which this
experiment could be performed. That way, and as required, the hardware
and cloud cost for this experiment were zero. Experiments were
announced through a post in Twitter and other social networks, and
results were published here \cite{DBLP:conf/gecco/GuervosG15}. For the
purpose of this paper, we repeated the announcement several times
through the months of April and then by the beginning of August. All
in all, we have the set of runs with the characteristics shown in
Table \ref{tab:runs}. In general, every experiment took several
days. No particular care was taken about the time of the announcement
or the particular wording. Every {\em experiment} consisted in running
until the solution of the 40-trap problem was found. When the correct
solution was sent to the server, the counter was updated and the pool
of solutions reset to the void set. No special care was taken to wait
until all clients had finished, thus it might happen that, in fact,
the islands running in the browser {\em spill} from one experiment to
the next. However, previous experiments have proved that the influence
of these islands in the next experiment is, in fact negligible.
%
\begin{table*}d
\caption{Characteristics of the volunteer computing initial runs. \label{tab:runs}}
\begin{center}
\begin{tabular}{l|rr}
\hline
Date & \# experiments & Different IPs \\
\hline 
April 4th 4/4 & 57 & 191 \\
April 24th 4/24 &  231 & 559 \\
July 31th 7/31 & 97 & 179 \\
\hline 
\end{tabular}
\end{center}
\end{table*}
%
\begin{table*}
\caption{Summary of time per run, number of IPS and number of {\tt
    PUT}s per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline 
4/4 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 &  5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline 
\end{tabular}
\end{center}
\end{table*}
%
The table shows that every run included more than 50 experiments. The
number of different IPs intervening in them varied from more than one
hundred to more than five hundred in the second experiment.

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}. This table shows the median number of IPs
intervening in each experiment,  median of the time needed
to finish the experiment, median number of HTTP {\tt PUT}s per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
found previously \cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same ballpark and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the last two columns, that measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: this experiment takes
place in a browser and in fact most of the time is devoted to
presenting the charts; the server is not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
fair is fair, and we were looking for creating a framework that
allowed to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results is that although it will
happen, one out of roughly 5 or roughly 20, depending on what baseline
experiment you choose, it will not happen always or even a majority of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found; this is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. However, it puts a probably
unnecesary cap on the time, or number of experiments, that an user
will voluntarily perform.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{img/time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots. From left to right, experiments 4/4, 4/24 and 7/31. Shades
of blue indicate how many experiments included that many unique IPs.}
\label{fig:duration}
\end{figure}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{img/time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs rank, with $y$ axis in a
  logarithmic scale. Dot color is related to the number of IPs
  participating in the experiment. From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:zipf:os}
\end{figure}
%
We will have to analyze experiment data a bit further to find out why
this happens and also if there is some pattern in the three sets of
experiments. An interesting question to ask, for instance, is if
adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, {\em adding} more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
wit experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more so taking
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
with a bigger size to check this hypothesis; however, at this point we
have not found this convenient since there are several other issues
that will be solved.

@@

It is interesting also to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, and of the order of
10 seconds. As it can be seen, it drops sharply implying there are
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP {\tt PUT}s, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of {\tt PUT}s per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline 
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\  
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\    
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\   
\hline 
\end{tabular}
\end{center}
\end{table*}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and  scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, negative towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiment share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
\ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

At any rate, this also shows that a convenient way of increasing the
computing power would be increasing this minimum amount per user. This
issue led us to create the next version of the framework, which is
presented next.

@@

In order to improve the number of cycles per user, we improved the
basic architecture in several different aspects. By using WebWorkers
the algorithm kept running on the background even if the {\sf NodIO}
tab was not in focus. It also allowed to run several clients per
browser, as opposed to a single one as we did previously. Another
improvement involved restart of the client once the solution was
found, so that as long as the user was visiting the page it was
running an evolutionary algorithm. Finally, a small algorithmic
improvement was included: instead of having a fixed population,
population size was random and different for each client.

JavaScript has commonly been used to develop client-side scripts
that handle the user interface, communicate asynchronously with servers and 
update the content that is displayed \cite{flanagan2006javascript}.
Long-running scripts like those proposed in this work can interfere with the
responsiveness of the web application, because everything is handled by a 
single thread. The Web Workers specification \cite{hickson2012web} defines an 
API that allows Web developers to execute long-running scripts in parallel
with the scripts of the main page. Worker scripts are spawned in the 
background allowing a thread-like operation using message-passing as the 
communication mechanism between the worker's execution environment and the 
main thread. According to the Web Workers W3C Candidate Recomendation 
\cite{hickson2012web} workers are expected to be long-lived, they have a high 
start-up performance cost, and a high per-instance memory cost. The Web Worker
API is implemented in current versions of both desktop and mobile web browsers.
For volunteer computing using the Web Worker API brings several advantages 
over a single threaded implementation:

\begin{itemize}
\item If the browser uses a tabbed document interface, the worker script 
keeps running in the background even it the user brings forward another tabbed
window.
\item Several instances of NodEO can be executed in parallel in a single web
page. 
\item Many immplemantations of the Web Worker API can use multiple CPUs for
the execution of worker scripts, in particular the execution in multiple CPUs
was tested in desktop verions of Chrome, FireFox and Safari.
\item Each NodEO instance can be restarted independenly.
\end{itemize}

A sequence diagram of the W$^2$ version of {\sf NodIO} is presented 
in Figure~\ref{fig:system:w2}. In this version, clients use the Web Workers
API to run two NodEO instances per browser. The sequence diagram shows the 
interaction between processess in a time sequence. In this system there are
two kinds of asynchronus messages: HTTP Requests shown as bold arrows between 
the client and the server where PUT and GET HTTP Methods are represented by 
red and blue arrows respectively. There are also messages between workers and 
the main thread in locally in the client, this are shown as thin solid arrows.
There are three processes involved:
\begin{itemize}
\item {\em nodejs} This is the server side process, implented in the nodejs 
framework. Web request handlers are implemented using the Express web 
framework. For keeping a log of the execution the Winston library is used, and 
finally the NodeIO framework is responsible for keeping the shared pool.
\item {\em Client main thread}  This process handles the user interaction
using the JQuery framework. The Chart.js library is used to display a chart of
the best fitness in the population over time. The main script also creates 
worker instances and handles the user interface.
\item {\em Worker global scope} The worker execution space does not have 
access to the Document Object Model (DOM) or JQuery objects in the main thread.
Asynchronus HTTP requests to the server are implemented using the XHR object. 
\end{itemize}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=4in]{img/Algorithm.pdf}
\caption{Description of the W$^2$ version of {\sf NodIO}. In this
  version, clients use Web Workers to run the evolutionary algorithm,
  with two of them per browser.}  
\label{fig:system:w2}
\end{figure}

The sequence of events is as follows:
\begin{enumerate}
\item First the {\sf NodIO} process is started in nodejs. As part of the 
initialization procedure the number of experiment is initialised and a log 
file is created. The shared pool implemented as an array is reset.
\item A volunteer follows the link of the experiment triggering a GET HTTP 
request and the page is loaded.
\item The main script renders the page and creates the worker instance. In 
order to create a dedicated worker instance another GET request is sent to the 
server in order to recieve the worker implementation script. 
Each worker instance executes one or more evolutionary algorithms 
{\em islands}. Each starts with random population and it is assigned a 
universally unique  identifier (UUID). The UUID is included in each of the HTTP
requests sent to the server.  
The initial paramaters for each island
can be distinct. Once the worker is created and the parameters have been set,
a message is sent to the worker to start the algorithm.
\item The worker first creates the population following with the previously 
specified parameters. Once the population has been initialised, the EA 
generation loop is started. After n generations the state of the population 
is evaluated. First the chromosome with the best fitness is sent to the shared
population with a PUT HTTP request. If the best fitness is not a global best,
the EA loop must continue, but first a chromosome is randomly selected from 
the shared pupulation. Also the current generation number and the best 
chromosome is sent to the client main thread in order to update the DOM to 
display the current state of the island (described next). These messages are
all asynchronous so even before a message is recieved by target the EA loop of
the worker could continue. If the best fitness is a global best the best 
chromosme is sent to the server and the loop ends. A message is sent to the 
main thread indicating that the best individual has been found.
\item The main thread recieves messages from workers as events handled by 
a callback function. When an iteration message is recived ,the page is 
updated by displaying a dinamic plot of the geneartion number and best 
fitness, and a representation of the chromosme is also generated.   
\item When a global best is recieved from an island the current experiment 
ends, the experiment number is incremented and the population array is reset.  
\item When the main thread recieves a message the a worker has ended the EA
loop, that worker is reinitialized. The worker process is not ended, to
avoid the cost, only the parameters and population is reset and a new UIID 
assigned.
\end{enumerate}
%
\begin{table*}[!htb]
\caption{Summary of {\sf NodIO-W$^2$} and comparison with the previous
  experiments, which have been aggregated to compute central measures. \label{tab:summary:ww}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline 
{\sf NodIO} & 5 & 29 & 123 & 14 & 37.43\% & 3.40\% & 0.10 \\
{\sf NodIO-W$^2$} & 4  & 16 & 7.36 & 40 & 89\% & 36.90\% & 0.4336061 \\
\hline 
\end{tabular}
\end{center}
\end{table*}
%

The experiments were run in the same way as before, using several
Twitter announcements throughout several days by the end of July
2015. Eventually, more than one thousand experiments were completed in
a matter of days. A summary of results are shown in Table
\ref{tab:summary:ww}, comparing it with the aggregate of the results
obtained with the initial version of {\sf NodIO}. These results are
remarkably different, being similar only in the median and maximum
number of IPs, although in this case it would be combination
IP-worker, being considered every worker a different IP. The median
number of seconds is almost 80\% better than the previous version, and
the median number of PUTs is three times better; this causes that most
of experiment take less time than one of the baselines, and one third
less than the second baseline. Thus, the conclusion in this case is
that we can obtain, through volunteer computation, a result that is,
most of the times, better than we would using a similar, non-parallel,
desktop setup, which makes this system suitable for massively
distributed evolutionary algorithms.

However, it is interesting to note why this is so, and the first hint
is the inter-experiment correlation between the number of IPs in
succesive experiments. While before it was an unremarkable 10\%, it is
now more than 43\%, making the numbef of IPs in contiguous experiments
highly correlated. The main reason for this is the setup in {\sf
  NodIO-W$^2$}, which restarts an island after a solution has been
found and also is running even if the tab is not on the
foreground. This means that volunteers can leave the experiment
running for as long as they want and they will be contributing to
experiment after experiment, unlike the previous version, where they
stopped contributing after the solution was found. This is also
reflected in the number of PUTs, which is the number of generations
divided by 100; more than 50\% of the volunteers run for more than
4000 generations. All things considered, this means than there are
many more computers running {\em simultaneously}, leading to this
almost 5-fold increase in running time.
